
\begin{thebibliography}{99}
    \rhead[]{Bibliografía}
	
	\addcontentsline{toc}{chapter}{Bibliografía}

\bibitem{Abdi 2007} \textsc{Abdi, H.,} \emph{The method of least squares.} Encyclopedia of measurement and statistics, 1, 530-532, 2007.


\bibitem{Abdi 2010} \textsc{Abdi, H., Williams, L. J.: } Principal component analysis. \emph{Wiley Interdisciplinary Reviews: Computational Statistics}  2(4), 433–459, 2010.


\bibitem{Galindo 2015}\textsc{Andrade-Sánchez, A. I., Galindo-Villardón, M. P., y Cuevas Romo, J. } \emph{Análisis multivariante del perfil psicológico de los deportistas universitarios: Aplicación del CPRD en México}. Educación Física y Ciencia, 17(2), 2015.  Recuperado de \url{http://www.efyc.fahce.unlp.edu.ar/article/view/EFyCv17n02a02/}

%\bibitem{Batista, 1989}
%\textbf{Batista Foguet, J.M. y Martínez Arias, R. (1989)}\textit{Análisis Multivariante: Análisis en Componentes Principales. } Editorial Hispano Europea.

\bibitem{Biau 2016}\textsc{Biau, G.; Scornet, E}., \textit{A random forest guided tour}. Test, 25, 197-227, 2016.

\bibitem{Breiman 1984}\textsc{Breiman, L., Friedman, J., Stone, C. J., Olshen, R. A.} \emph{Classification and regression trees. }Chapman \& Hall, New York, London, 1984.

\bibitem{Breiman 1996}\textsc{Breiman, L. } \emph{Bagging predictors.} Machine learning, 24, 123-140, 1996.

\bibitem{Breiman 2001}\textsc{Breiman, L. } \emph{Random forests}. Machine learning, 45, 5-32, 2001.

\bibitem{Breiman 2004a}\textsc{Breiman, L. } \emph{Consistency for a simple model of random forests}. University of California at Berkeley. Technical Report, 670, 2004.

\bibitem{Brown 2004} \textsc{Brown, S. D.,  Myles, A. J., Feudale, R. N., Liu, Y., and Woody, N. A.,} \emph{An introduction to decision tree modeling}. Journal of Chemometrics: A Journal of the Chemometrics Society, 18(6), 275-285, 2004.

\bibitem{Chatfield 1989} \textsc{Chatfield, C. and Collins A.J}. {\em Introduction to multivariate analysis}, Chapman and Hall, London, 1989.

\bibitem{Cuadras 2014} \textsc{Cuadras, C.M. }, \textit{Nuevos métodos de Análisis Multivariante}, CMC Editions, Barcelona, 
2014.

\bibitem{Diaz 2006}\textsc{Díaz-Uriarte, R., Alvarez de Andrés, S. }\emph{ Gene selection and classification of microarray data using random forest.} BMC bioinformatics, 7, 1-13, 2006.

\bibitem{Diez 2002}\textsc{Diez, J. P., Hernandis, S. P., Rouma, A. C. } \emph{El consumo de inhalables y cannabis en la preadolescencia: Análisis multivariado de factores predisponentes}. Anales de Psicología/Annals of Psychology, 18(1), 77-93, 2002.

\bibitem{Divakaran 2022}\textsc{Divakaran, S. . }Data Science: Principles and Concepts in Modeling Decision Trees.\emph{ Data Science in Agriculture and Natural Resource Management}, 55-74, 2022.

\bibitem{Alubias} \textsc{Dry Bean Dataset}. UCI Machine Learning Repository.  \url{https://doi.org/10.24432/C50S4B.}, 2020.

\bibitem{Echeverri 2015} \textsc{Machado-Duque, M. E., Echeverri Chabur, J. E., y Machado-Alba, J. E. } \emph{Somnolencia diurna excesiva, mala calidad del sueño y bajo rendimiento académico en estudiantes de Medicina.} Revista colombiana de Psiquiatría, 44(3), 137-142, 2015.


\bibitem{Eckart 1936}\textsc{Eckart, C., Young, G. } \emph{The approximation of one matrix by another of lower rank.} Psychometrika, 1(3), 211-218, 1936.

\bibitem{Ensum 2005}\textsc{Ensum, J., Pollard, R., \& Taylor, S. (2005).}\emph{ Applications of logistic regression to shots at goal at association football.} In Science and football V: the proceedings of the Fifth World Congress on Science and Football, 214-221, 2005.

\bibitem{Everitt 2002} \textsc{Everitt, B., Ismail, K., David, A. S., Wessely, S. }. \emph{Searching for a Gulf War syndrome using cluster analysis.} Psychological medicine, 32(8), 1371-1378, 2002.

\bibitem{Everitt 2011a}\textsc{Everitt, B., and Hothorn, T. (2011)}. \emph{An introduction to applied multivariate analysis with R}. Springer Science \& Business Media.

\bibitem{Everitt 2011}\textsc{Everitt, B. S., Leese, M., Stahl, D., \& Landau, S. (2011}). \emph{Cluster analysis}. John Wiley \& Sons, United Kingdom, 2011. 

\bibitem{Iris Fisher} \textsc{Fisher,R. A.. (1988).} \emph{Iris}. UCI Machine Learning Repository. \url{https://doi.org/10.24432/C56C76}, 1988.

\bibitem{Galton  1889} \textsc{Galton, F. } \emph{Co-relations and their measurement, chiefly from anthropometric data.} Proceedings of the Royal Society of London, 45(273-279), 135-145, 1889.

\bibitem{Golub 1987}\textsc{Golub, G. H., Hoffman, A., \& Stewart, G. W. }. A generalization of the Eckart-Young-Mirsky matrix approximation theorem. \emph{Linear Algebra and its applications}, 88, 317-327, 1987. 

\bibitem{Asari 2004} \textsc{Gottumukkal, R., \& Asari, V. K. } An improved face recognition technique based on modular PCA approach.\emph{ Pattern Recognition Letters,} 25(4), 429-436, 2004.

\bibitem{Greene 2008}\textsc{Greene,  W.H.}. \emph{Econometric analysis (6th ed.)} Pearson Prentice Hall, New Jersey, 2008.

\bibitem{Grossi 2007}\textsc{Grossi, E.,  Buscema, M.  }Introduction to artificial neural networks. \emph{European journal of gastroenterology \& hepatology,} 19(12), 1046-1054, 2007.

\bibitem{Hair 1995}\textsc{Hair, J. F., Black, W. C., Tatham, R. L., and Anderson, R. E. .}
\textit{ Multivariate data analysis with readings.} Prentice-Hall International, Englewood Cliffs, New Jersey, 1995.

\bibitem{Hartigan 1975}\textsc{Hartigan, J. A. }\emph{ Clustering algorithms}. John Wiley \& Sons, Inc, New York, 1975. 

\bibitem{Hastie 2001} \textsc{Hastie, T., Tibshirani, R. Friedman J.}, \textit{The Elements of Statistical Learning, Data Mining, Inference and Prediction} Springer, Stanford, 2001.

\bibitem{Hesterberg 2011}\textsc{Hesterberg, T.} \emph{Bootstrap}. Wiley Interdisciplinary Reviews: Computational Statistics, 3(6), 497-526, 2011.

\bibitem{Hochreiter 1997}\textsc{Hochreiter, S.,  Schmidhuber, J. } Long short-term memory.\emph{ Neural computation}, 9(8), 1735-1780, 1997.

\bibitem{Hornik 1989} \textsc{Hornik, K., Stinchcombe, M., and White, H. } Multilayer feedforward networks are universal approximators. \emph{Neural networks}, 2(5), 359-366, 1989.
\url{https://doi.org/10.1016/0893-6080(89)90020-8}

\bibitem{Hotelling 1933}\textsc{Hotelling H. } Analysis of a complex of statistical variables
into principal components. \emph{ Journal of educational psychology }
25:417–441, 1933.

\bibitem{Jain 1988} \textsc{Jain, A.K., Dubes R.C,} \textit{Algorithms for clustering data.} Prentice-Hall Inc, New Jersey,  1988.

\bibitem{James 2013}
\textsc{James G, Witten D, Hastie T, Tibshirani R.} \emph{An Introduction to Statistical Learning: With Applications in R.} New York, Springer; 2021. 

\bibitem{Johnson 2007}\textsc{Johnson, R. A., \& Wichern, D. W.}. \emph{Applied multivariate statistical analysis.(6th edition)} Pearson Prentice Hall, New Jersey, 2007.

\bibitem{Johnson 1963} \textsc{Johnson, R. M. } \emph{On a theorem stated by Eckart and Young.} Psychometrika, 28(3), 259-263,1963.

\bibitem{Jollife 1986} \textsc{Jollife I.T}. \emph{Principal Component Analysis}, Springer, New York, 1986.



\bibitem{Kass 1980}\textsc{Kass, G. V. }. \emph{An exploratory technique for investigating large quantities of categorical data.} Journal of the Royal Statistical Society: Series C (Applied Statistics), 29(2), 119-127, 1980.

%\bibitem{Koppen 2000} \textsc{Köppen, M. (2000)} The curse of dimensionality. \textit{5th online world conference on soft computing in industrial applications (WSC5)}  (Vol. 1, pp. 4-8).

\bibitem{Koklu 2020}\textsc{Koklu, M., \& Ozkan, I. A.} \emph{Multiclass classification of dry beans using computer vision and machine learning techniques.} Computers and Electronics in Agriculture, 174, 105507, 2020.

\bibitem{Lawless 2010} \textsc{Lawless, J. F., \& Yuan, Y.} \emph{Estimation of prediction error for survival models.} \emph{Statistics in medicine,} 29(2), 262-274, 2010.

\bibitem{Roberto 2008} \textsc{Lopez, R., Balsa-Canto, E. and Oñate, E. }  Neural networks for variational problems in engineering International Journal for Numerical Methods Engineering., 75 \url{https://doi.org/10.1002/nme.2304} 1341-1360, 2008.

\bibitem{Lebart 1984}\textsc{Lebart, L., Morineau, A.,  Warwick, K. M. }.\textit{Multivariate Descriptive Analysis: Correspondence analysis and related techniques for large matrices}. Wiley, New York, 1984.

\bibitem{Loh 2011}\textsc{Loh, W. Y. } \emph{Classification and regression trees.} Wiley interdisciplinary reviews: data mining and knowledge discovery, 1(1), 14-23, 2011.

\bibitem{Loh 2014}\textsc{Loh, W. Y. } \emph{Fifty years of classification and regression trees.} International Statistical Review, 82(3), 329-348, 2014.

\bibitem{Livingstone 1997}\textsc{Livingstone, D. J., Manallack, D. T., Tetko, I. V. } \emph{Data modelling with neural networks: Advantages and limitations.} Journal of computer-aided molecular design, 11, 135-142, 1997.

\bibitem{Mahesh 2020}\textsc{Mahesh, B. } \emph{Machine learning algorithms-a review.} International Journal of Science and Research (IJSR), 9, 381-386, 2020.

\bibitem{Mamidi 2021}\textsc{Mamidi, R. R. }. \emph{Application of Neural Networks for Prediction of Double Fed Induction Generator’s Equivalent Circuit Parameters used in Wind Generators.} IRJCS:: International Research Journal of Computer Science, Volume VIII, 23-31, 2021.

\bibitem{Mardia 1979}\textsc{Mardia, K.V., Kent, J.T. Bibby J.M.; }. \emph{Multivariate analysis}. Academic Press, San Diego , 1979.

\bibitem{Rosario 1999} 
\textsc{Martínez R.} \emph{El análisis multivariante en la investigación científica.} Madrid: La Muralla; 1999.

\bibitem{Matplotlib}\textsc{Matplotlib Documentation:} URL: \url{https://matplotlib.org/}. Consultado en varias fechas. 

\bibitem{Morrison 1976}\textsc{Morrison. D.}\textit{ Multivariate statistical methods (2nd ed)}. McGraw-Hill Kogakusha,(1976).

\bibitem{Nerini 2007}\textsc{Nerini, D.,  Ghattas, B. }\emph{ Classifying densities using functional regression trees: Applications in oceanology.} Computational Statistics \& Data Analysis 51(10), 4984-4993,(1976).

\bibitem{Neural Designer} \textsc{Neural Designer Blog} URL:\url{https://www.neuraldesigner.com/}, consultado en varias fechas.

\bibitem{Numpy}\textsc{Numpy Documentation} URL: \url{https://numpy.org/doc/stable/}. Consultado en varias fechas.

\bibitem{Okazaki 2006}\textsc{Okazaki, S. (2006).}\emph{ What do we know about mobile Internet adopters? A cluster analysis.} Information \& Management, 43(2), 127-141, 2006.

\bibitem{Pages 2005}\textsc{Pagès, J. } \emph{Collection and analysis of perceived product inter-distances using multiple factor analysis: Application to the study of 10 white wines from the Loire Valley.} Food quality and preference, 16(7), 642-649, 2005. DOI: \url{https://doi.org/10.1016/j.foodqual.2005.01.006
}

\bibitem{Pandas} \textsc{Pandas Documentation} URL:\url{https://pandas.pydata.org/docs/}. Consultado en varias fechas.

\bibitem{Peña 2002} 
\textsc{Peña D.} \emph{Análisis de datos multivariantes.} Madrid: McGraw-Hill; 2002.

\bibitem{Pearson 1901} \textsc{Pearson, K. }. LIII. On lines and planes of closest fit to systems of points in space. \emph{The London, Edinburgh, and Dublin philosophical magazine and journal of science,} 2(11), 559-572, 1901.

\bibitem{Quinlan 2014}\textsc{Quinlan, J. R. }. \emph{C4. 5: programs for machine learning.} Elsevier, 2014.

\bibitem{Rencher 2002} \textsc{Rencher, A. C.}. \emph{Methods of multivariate analysis.} New York: A John Wiley \& Sons. Inc. Publication, 2002.

\bibitem{Requena}
\textsc{A. Requena, R. Quintanilla, J.M. Bolarín, A. Vázquez, A. Bastida, J. Zúñiga y L.M. Tomás.}\emph{Nuevas Tecnologías y Contaminación de Atmósferas, para PYMEs.} VI-3-4, Pag. 3 Universidad de Murcia. Spain. \url{https://www.um.es/LEQ/Atmosferas/Ch-VI-3/F63s4p3.htm}.

\bibitem{Ringnér 2008}\textsc{Ringnér, M. (2008).}\emph{What is principal component analysis?}.Nature biotechnology, 26(3), 303-304.

\bibitem{Witten 2010}\textsc{Witten, D. M., \& Tibshirani, R. } \emph{Survival analysis with high-dimensional covariates}. \emph{Statistical methods in medical research,} 19(1), 29-51, 2010.

\bibitem{Scitovski 2021}\textsc{Scitovski, R., Sabo, K., Martínez-Álvarez, F., \& Ungar, Š. } \emph{Cluster analysis and Applications} Cham, Springer.

\bibitem{Scikit-Learn} \textsc{Scikit Learn Documentation} URL: \url{https://scikit-learn.org/stable}, consultado en varias fechas. 

\bibitem{Song 2015}\textsc{Song, Y. Y., and Ying, L. U. } \emph{Decision tree methods: applications for classification and prediction}. Shanghai archives of psychiatry, 27(2), 130, 2015.

\bibitem{Spearman 1904}\textsc{Spearman, C.} \emph{“General Intelligence,” Objectively Determined and Measured.} The American Journal of Psychology, 15(2), 201–292, 1904. DOI: \url{https://doi.org/10.2307/1412107}

\bibitem{TensorFlow}\textsc{Tensor Flow Documentation} URL: \url{https://www.tensorflow.org}. Consultado en varias fechas.

\bibitem{Vincent 1953}\textsc{Vincent, D. F. (1953).} \emph{The origin and development of factor analysis.} Journal of the Royal Statistical Society: Series C (Applied Statistics), 2(2), 107-117, 1953.


%\bibitem{González 2015} \textsc{González García, N., y Taborda Londoño, A.} \emph{Análisis de componentes principales Sparse: formulación, algoritmos e implicaciones en el análisis de datos.} Salamanca,  2015

\bibitem{Yeh 2007}\textsc{Yeh I. }. \emph{Concrete Compressive Strength.} UCI Machine Learning Repository, 2007. \url{https://doi.org/10.24432/C5PK67.}

\end{thebibliography}
 